---
title: "Analysis"
number-sections: false
toc: false
---

::: {.callout-note title="📌 Data Preparation"}
The dataset used in this project was obtained from the CDC Pregnancy Vaccination Surveillance System, which provides estimates of maternal vaccination coverage in the United States. The data were downloaded as a CSV file and imported into R for cleaning and analysis.

To prepare the dataset, variable names were standardized to lowercase and formatted for clarity (e.g., *Survey Year/Influenza Season* → **year**). Coverage estimates were converted to numeric values, and confidence interval fields were preserved in their original text form. The dataset includes information on vaccine type (Influenza or Tdap), geographic unit (national or state), survey year, demographic subgroup, and estimated vaccination coverage.

No unformatted console output (such as `head()`, `str()`, or raw print statements) is included in this report. Instead, dataset characteristics are summarized using formatted tables to ensure readability and professional presentation.
:::

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(dplyr)
library(ggprism)
library(sf)
```

```{r message=FALSE, warning=FALSE}
vacc_raw <- read_csv("~/Desktop/Vaccination_Coverage_among_Pregnant_Women_20251202.csv")
```

```{r message=FALSE, warning=FALSE}
colnames(vacc_raw)
```

```{r}
 vacc_clean <- vacc_raw |>
  dplyr::rename(
    vaccine         = Vaccine,
    geography_type  = `Geography Type`,
    geography       = Geography,
    year            = `Survey Year/Influenza Season`,
    dimension_type  = `Dimension Type`,
    dimension       = Dimension,
    estimate        = `Estimate (%)`,
    ci              = `95% CI (%)`,
    sample_size     = `Sample Size`
  ) |>
  mutate(
    year        = as.integer(year),
    estimate    = as.numeric(estimate),
    sample_size = as.integer(sample_size),
    vaccine        = factor(vaccine),
    geography_type = factor(geography_type),
    dimension_type = factor(dimension_type)
  ) |>
  filter(
    !is.na(estimate),
    !is.na(sample_size),
    !is.na(year)
  )

```

```{r message=FALSE, warning=FALSE}
dataset_summary <- vacc_clean |>
  summarise(
    total_rows        = n(),
    min_year          = min(year),
    max_year          = max(year),
    n_vaccines        = n_distinct(vaccine),
    n_geographies     = n_distinct(geography),
    n_dimension_types = n_distinct(dimension_type)
  )

knitr::kable(dataset_summary, caption = "Summary of Dataset Characteristics")


```

```{r message=FALSE, warning=FALSE}
data_overview <- tibble(
  `Number of Rows` = nrow(vacc_clean),
  `Number of Columns` = ncol(vacc_clean)
)

kable(data_overview, caption = "Dataset Size: Rows and Columns")
```

```{r message=FALSE, warning=FALSE}
dimension_table <- vacc_clean |>
  count(dimension_type, sort = TRUE)
kable(dimension_table, caption = "Number of Observations by Dimension Type")

```

```{r message=FALSE, warning=FALSE}
variable_list <- tibble(
  Variable = names(vacc_clean),
  Description = c(
    "Vaccine type (Influenza or Tdap)",
    "Geography level (National or State)",
    "State name or United States total",
    "Survey year (numeric)",
    "Type of demographic subgroup (Age or Race and Ethnicity)",
    "Specific subgroup (e.g., 18–24, Hispanic, Non-Hispanic White)",
    "Vaccination coverage estimate (%)",
    "Reported 95% confidence interval for the estimate (as a text range)",
    "Sample size used to compute the estimate"
  )
)

kable(variable_list, caption = "Variables Included in the Cleaned Dataset")
```

```{r message=FALSE, warning=FALSE}
na_summary <- vacc_clean |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  tidyr::pivot_longer(cols = everything(),
                      names_to = "Variable",
                      values_to = "Missing Values")

knitr::kable(na_summary, caption = "Number of Missing Values per Variable")

```

```{r message=FALSE, warning=FALSE}
vaccine_table <- vacc_clean |>
  count(vaccine, sort = TRUE)

kable(vaccine_table, caption = "Number of Observations by Vaccine Type")
```

❓Distribution of State-Level Coverage: Influenza

```{r message=FALSE, warning=FALSE}
vacc_state <- vacc_clean %>%
  filter(geography_type == "States")
infl_state <- vacc_state %>%
filter(vaccine == "Influenza")

ggplot(infl_state, aes(x = estimate)) +
geom_histogram(
bins = 20,
fill = "#4E79A7",
color = "white",
linewidth = 0.2,
alpha = 0.9
) +
geom_vline(
xintercept = mean(infl_state$estimate, na.rm = TRUE),
linetype = "dashed",
linewidth = 0.7
) +
labs(
title = "Distribution of State-Level Influenza Vaccination Coverage",
x = "Coverage (%)",
y = "Number of States"
) +
theme_prism(base_size = 12) +
theme(
plot.title = element_text(size = 14, face = "bold")
)

```

✅State-level influenza vaccination coverage among pregnant women shows a fairly wide distribution across the dataset, with most observations falling in the mid-range.

❓Tdap:

```{r}
tdap_state <- vacc_state %>%
filter(vaccine == "Tdap")

ggplot(tdap_state, aes(x = estimate)) +
geom_histogram(
bins = 20,
fill = "#E15759", 
color = "white",
linewidth = 0.2,
alpha = 0.9
) +
geom_vline(
xintercept = mean(tdap_state$estimate, na.rm = TRUE),
linetype = "dashed",
linewidth = 0.7
) +
labs(
title = "Distribution of State-Level Tdap Vaccination Coverage",
x = "Coverage (%)",
y = "Number of States"
) +
theme_prism(base_size = 12) +
theme(
plot.title = element_text(size = 14, face = "bold")
)
```

✅In contrast, Tdap vaccination coverage is generally shifted toward higher levels and appears more tightly clustered. The Tdap histogram shows a concentration of observations at relatively high coverage percentages, with fewer state–year combinations in the low-coverage range.

❓How do Influenza and Tdap coverage compare?

```{r message=FALSE, warning=FALSE}

  ggplot(vacc_clean, aes(x = vaccine, y = estimate, fill = vaccine)) +
  geom_boxplot(
    outlier.alpha = 0.20, 
    outlier.size  = 1,
    alpha         = 0.9
  ) +
  labs(
    title = "Influenza vs Tdap Vaccination Coverage",
    x = "Vaccine Type",
    y = "Coverage (%)"
  ) +
  scale_fill_prism(palette = "floral") +
  theme_prism(base_size = 14) +
  theme(
    legend.position = "none",
    plot.title      = element_text(size = 18, face = "bold"),
    axis.title.x    = element_text(size = 14),
    axis.title.y    = element_text(size = 14),
    axis.text       = element_text(size = 12)
  )

```

✅Tdap median coverage: higher. Influenza: lower and more variable

❓Vaccination Coverage Over Time

```{r message=FALSE, warning=FALSE}
trend <- vacc_clean %>%
  group_by(year) %>%
  summarise(mean_coverage = mean(estimate, na.rm = TRUE))

ggplot(trend, aes(x = year, y = mean_coverage)) +
  geom_line(color = "#2C73D2", linewidth = 1.6) +
  geom_point(color = "#FFA500", size = 3.5) +
  scale_y_continuous(limits = c(40, NA)) +
  scale_x_continuous(breaks = seq(min(trend$year), max(trend$year), 1)) +
  labs(
    title = "Maternal Vaccination Coverage Over Time",
    x = "Survey Year",
    y = "Mean Coverage (%)"
  ) +
  theme_prism(base_size = 14) +
  theme(plot.title = element_text(size = 18, face = "bold"))



```

✅Maternal vaccination coverage shows a gradual upward trend from 2012 to around 2019, followed by slight fluctuation in later years (2020–2021). This pattern reflects changing vaccine recommendations, improvements in outreach, and disruptions during the pandemic period.

```{r message=FALSE, warning=FALSE}
trend_data <- vacc_clean %>%
  filter(
    geography_type == "States",
    vaccine %in% c("Influenza", "Tdap")
  ) %>%

  mutate(year4 = stringr::str_extract(year, "^[0-9]{4}")) %>%
  filter(!is.na(year4),
         year4 >= "2012",
         year4 <= "2022") %>%
  group_by(year4, vaccine) %>%
  summarise(
    mean_coverage = mean(estimate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    year = factor(year4, levels = as.character(2012:2022))
  )

ggplot(trend_data, aes(x = year, y = mean_coverage, color = vaccine, group = vaccine)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 2.8) +
  scale_color_prism(palette = "floral") +
  labs(
    title = "Vaccination Trends Over Time (2012–2022)",
    x = "Year",
    y = "Average Coverage (%)",
    color = "Vaccine"
  ) +
  scale_x_discrete(drop = FALSE) +
  theme_prism(base_size = 12) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    plot.title   = element_text(size = 15, face = "bold")
  )
```

✅Tdap Coverage Increased Rapidly and Reached High, Stable Levels and Influenza Coverage Is Lower, More Variable, and Plateaued.

❓How do vaccination rates differ across demographic groups (age & race/ethnicity)?

Age:

```{r}
age_summary <- vacc_clean %>%
  filter(dimension_type == "Age",
         vaccine %in% c("Influenza", "Tdap")) %>%
  group_by(vaccine, dimension) %>%
  summarise(
    mean_coverage   = mean(estimate, na.rm = TRUE),
    median_coverage = median(estimate, na.rm = TRUE),
    n               = n()
  )

kable(age_summary, digits = 1,
      caption = "Vaccination Coverage by Age Group and Vaccine")
```

```{r message=FALSE, warning=FALSE}
age_data <- vacc_clean %>% filter(dimension_type == "Age")

ggplot(age_data, aes(x = dimension, y = estimate)) +
  geom_boxplot(fill = "#00A9A5", alpha = 0.85) +
  labs(
    title = "Vaccination Coverage Across Age Groups",
    x = "Age Group",
    y = "Coverage (%)"
  ) +
  theme_prism(base_size = 14) +
  theme(plot.title = element_text(size = 18, face = "bold"))

```

✅Coverage is lowest among the youngest pregnant individuals (18–24) and gradually increases with age. Older groups (30–39) consistently show higher uptake.

❓Race & Ethnicity:

```{r message=FALSE, warning=FALSE}
race_summary <- vacc_clean %>%
  filter(dimension_type == "Race and Ethnicity",
         vaccine %in% c("Influenza", "Tdap")) %>%
  group_by(vaccine, dimension) %>%
  summarise(
    mean_coverage   = mean(estimate, na.rm = TRUE),
    median_coverage = median(estimate, na.rm = TRUE),
    n               = n()
  )

kable(race_summary, digits = 1,
      caption = "Vaccination Coverage by Race/Ethnicity and Vaccine")
```

```{r message=FALSE, warning=FALSE}
race_data <- vacc_clean %>%
  filter(dimension_type == "Race and Ethnicity") %>%
  group_by(dimension) %>%
  summarise(mean_coverage = mean(estimate, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(
    dimension_label = dplyr::recode(
      dimension,
      "Other or Multiple Races, Non-Hispanic" = "Other or Multiple Races,\nNon-Hispanic",
      "White, Non-Hispanic"                  = "White,\nNon-Hispanic",
      "Black, Non-Hispanic"                  = "Black,\nNon-Hispanic",
      .default = dimension
    )
  )

ggplot(race_data,
       aes(x = reorder(dimension_label, mean_coverage),
           y = mean_coverage,
           fill = dimension_label)) +
  geom_col(width = 0.65, alpha = 0.9) +
  coord_flip() +
  labs(
    title = "Vaccination Coverage by Race/Ethnicity",
    x = "",
    y = "Mean Coverage (%)"
  ) +
  scale_fill_prism(palette = "floral") +
  theme_prism(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title  = element_text(size = 14, face = "bold"),
    axis.text.y = element_text(size = 11),
    axis.text.x = element_text(size = 11)
  )

```

✅The differences between the groups are not very clear. Hispanic and White (Non-Hispanic) have similar uptake. Black group shows the lowest among all race.

❓Influenza vs Tdap vaccination coverage over states:

```{r message=FALSE, warning=FALSE}

get_state_map <- function(vax_name) {
  vacc_clean %>%
    filter(
      geography_type == "States",
      vaccine == vax_name,
      year == 2022
    ) %>%
    group_by(geography) %>%
    slice_max(sample_size, n = 1) %>%
    ungroup() %>%
    transmute(
      state_name     = tolower(geography),
      estimate_state = estimate
    )
}

influenza_states <- get_state_map("Influenza")
tdap_states      <- get_state_map("Tdap")

states_geo <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE)) %>%
  mutate(state_name = tolower(ID))

combined_states <- bind_rows(
  influenza_states %>% mutate(vaccine = "Influenza"),
  tdap_states      %>% mutate(vaccine = "Tdap")
)

combined_map <- states_geo %>%
  left_join(combined_states, by = "state_name")

ggplot(combined_map) +
  geom_sf(
    aes(fill = estimate_state),
    color = "white",
    size = 0.2
  ) +
  scale_fill_gradient(
    low = "#FDE68A",
    high = "#B45309",
    na.value = "lightgray",   
    name = "Vaccination (%)"
  ) +
  facet_wrap(~ vaccine, nrow = 1) +
  labs(
    title    = "Influenza vs Tdap Vaccination Coverage (2022)",
    subtitle = "Pregnant Women • Missing states shown in gray",
    caption  = "Source: CDC Pregnancy Vaccination Surveillance System"
  ) +
  theme_void(base_size = 14) +
  theme(
    strip.text    = element_text(size = 15, face = "bold"),
    plot.title    = element_text(size = 22, face = "bold"),
    plot.subtitle = element_text(size = 14)
  )

```

✅Because the PRAMS Internet Panel Survey relies on voluntary state participation, not all 50 states contribute data each year. State participation changes over time, and in 2022 the dataset is missing 22 states. This incomplete geographic coverage makes it difficult to assess true national vaccination levels and introduces uncertainty when comparing trends or geographic variation. 2022 is used because its the lastest year from the dataset. Also, each year from the dataset has missing states so this dataset limits the ability to see the true nationwide vaccination coverage, so only 2022 is shown.

```{r message=FALSE, warning=FALSE}
all_states <- state.name

missing_by_year <- vacc_state %>%
  filter(!is.na(year)) %>%
  distinct(year, geography) %>%
  group_by(year) %>%
  summarise(
    present_states = list(unique(geography)),
    missing_states = list(setdiff(all_states, present_states[[1]])),
    n_missing = length(missing_states[[1]])
  )

missing_by_year

```

These are the missing states from the dataset.

```{r}
year_vax_summary <- vacc_clean %>%
  filter(!is.na(year), !is.na(vaccine)) %>%
  mutate(year4 = stringr::str_extract(year, "^[0-9]{4}")) %>%
  filter(
    !is.na(year4),
    year4 >= "2012",
    year4 <= "2022"
  ) %>%
  group_by(year4, vaccine) %>%
  summarise(
    coverage = weighted.mean(estimate, sample_size, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    year = factor(year4, levels = as.character(2012:2022))
  )

print(year_vax_summary)

ggplot(year_vax_summary, aes(x = year, y = vaccine, fill = coverage)) +
  geom_tile(color = "white") +
  scale_fill_gradient(
    low  = "#FDE68A",
    high = "#B45309",
    name = "Coverage (%)"
  ) +
  labs(
    title    = "Maternal Vaccination Coverage by Year and Vaccine",
    x        = "Year",
    y        = "Vaccine",
    caption  = "Values are sample-size–weighted mean estimates."
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    plot.title   = element_text(size = 18, face = "bold")
  )
```

✅The heatmap shows that Tdap vaccination coverage increased rapidly from 2012 to 2016 and remained high through 2022, while influenza coverage rose more slowly and fluctuated, with a decline after 2020. Overall, Tdap consistently demonstrated higher uptake than influenza among pregnant women.
